---
title: "Analysis of Traffic Offences Across Toronto Neighborhoods"
subtitle: "The effect of neighbourhoods incomes"
author: 
  - Gadiel David Flores
thanks: "Code and data are available at: https://open.toronto.ca/dataset/police-annual-statistical-report-tickets-issued/"
date: today
date-format: long
abstract: "In this paper, we will be looking at police annual reports of tickets isssued and analayzing the frequency in tickets issued based on different neighbourhoods. We will focus on which neighbourhoods exhibit the highest number of specific traffic offences (e.g., Aggressive Driving, Speeding). and are there observable geographical patterns or clusters of these offences based on neighbourhood? Additionally, how do the averages relate to the overall frequency of offences across different neighbourhoods?"
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
```


# Introduction

Higher police presence could lead to higher traffic violations.
It is probable that lower income neighborhoods have a higher police presence and would lead to higher citations overall. Another pattern that may be noticed is more dangerous roads that are prone for human error that leads to more violations given and this could be a potential exploit by officers as they may know this information beforehand. We use @citeR and @TorontoPoliceData2024.

The remainder of this paper is structured as follows. @sec-data will explore our data from @TorontoPoliceData2024 and show any potential clusters and patterns that may be in the number of offences given, locations, and type of citation and we will use this to further analyze in @sec-model. Lastly, we will discuss our findings in @sec-result.

# Data {#sec-data}
In this section we will be specifically looking at the names of neighbourhoods, and how many citations they got from years [TODO: find the minimum year and the max year]. The data of the frequency of each citation can be seen (@fig-citations), and was extract from @TorontoPoliceData2024.

```{r}
#| label: fig-citations
#| fig-cap: Frequency of Citations by Neighbourhood
#| echo: false

# Load necessary libraries
library(ggplot2)
library(here)

# Load the data from the CSV file using 'here'
tickets_data <- read.csv(here::here("data/sim_data/tickets_issued_data.csv"))

# bar plot focusing on neighbourhoods
ggplot(tickets_data, aes(x = neighbourhood)) +
  geom_bar() +
  labs(title = "Frequency of Citations by Neighbourhood", x = "Neighbourhood", y = "Count of Citations") +
  theme_minimal()

```
As it can be seen in this graph, [TODO: Talk about any patterns or observations from the real dataset]. In (@fig-income) we can see which neighborhoods have a higher income and which ones are lower income.This data came from the latest census of toronto and the only variables taken out from this data was the neighbourhoods and their average income. This data comes from [TODO: add reference to toronto data].

```{r}
#| label: fig-income
#| fig-cap: Toronto Neighbourhoods income
#| echo: false
#| warning: false
#| message: false

# Load necessary libraries
library(ggplot2)
library(here)

# Load the data from the CSV file using 'here'
income_data <- read.csv(here::here("data/sim_data/average_income.csv"))

# bar plot for average income by neighbourhood
ggplot(income_data, aes(x = neighbourhood, y = average_income)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Income by Neighbourhood", x = "Neighbourhood", y = "Average Income (in thousands)") +
  theme_minimal()


```

This data shows actual numbers of citations given out in each neighbourhood and the average income of each neighbourhood of each household comebined. Based on these graphs, we can see [TODO: Talk about any patterns or observations from the real dataset]. This raises a question on whether there is a correlation on income and the amount of citations that are given out. In the next seciton we will come with a model that can help us find a correlation. 


# Model {#sec-model}

The goal of our modeling strategy is twofold. First, to determine whether there is a relationship between neighborhood income and the number of offenses. Second, to assess how strong this relationship is.  
We will use a simple linear regression model to examine the effect of income on the number of citations issued in each neighborhood. Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model Setup

Define \( y_i \) as the number of offenses in a given neighborhood, and \( x_i \) as the average income of that neighborhood measured in thousands.

\[
E(Y|X) = \beta_0 + \beta_1(X)
\]

### Model Justification

A linear regression model was chosen because it allows us to analyze the effect of income on the number of citations. As neighborhood income increases or decreases, do offenses decrease or increase? This will help us evaluate the strength of the relationship between citations and average neighborhood income. While we could add more variables for a deeper analysis, that would fall outside the scope of this paper, which is focused solely on determining if a correlation exists between income and offenses.

# Results {#sec-results}

The linear regression model was applied to examine the relationship between neighborhood income and the number of traffic offenses. The model's equation is as follows:

\[
E(Y|X) = \beta_0 + \beta_1 \cdot \text{average\_income}
\]

where \( Y \) is the number of tickets issued, and \( X \) is the average income in thousands. The summary of the linear regression model is provided below in @ressidual_plots.

We expect a negative relationship between the number of offenses and neighborhood income, meaning that as income decreases, offenses increase. Our results are summarized in @ressidual_plots. Residuals vs. Predictors, Residuals vs. Fitted Values, and Normal Q-Q plots are used to verify that the four key assumptions of linear regression are not violated and that the model fits the data appropriately. We will assess whether linearity [TODO: check if linearity is violated], uncorrelated errors [TODO: check if uncorrelated errors are violated], constant variance [TODO: check for homoscedasticity], and normality of errors [TODO: check for normality] hold. Given these assumptions, we will [TODO: assess whether the results of the linear regression can be trusted].

# Discussion

## First Discussion Point {#sec-first-point}
To begin, we should focus on the main purpose of this paper, which is to determine whether there is a relationship between the number of offenses and the income of a neighborhood. As we can see from the results, [TODO: Discuss model results].

## Second Discussion Point
This model only considers two variables: the income and the number of offenses per neighborhood. It does not account for all possible factors that may influence the number of citations a neighborhood receives. For example, busier roads are more likely to have more offenses than quieter neighborhoods. Future analysis would need to examine these factors in more detail by incorporating variables such as road safety data, traffic volume, or demographic factors.

## Third Discussion Point
[TODO: Discuss how this information can help the public].

## Weaknesses and Next Steps

There may be bias in the data, as the income of a neighborhood could influence the amount of police activity in that area. This would lead to more citations being issued in lower-income neighborhoods due to higher police presence. Additionally, the linear regression model we used makes certain assumptions, which are further discussed in @sec-model-details, regarding how well the model fits the data. 

The next steps for this research would be to incorporate additional variables to provide a more comprehensive picture of the factors influencing the number of citations a neighborhood receives. Furthermore, using more sophisticated models could yield a more accurate understanding of the relationship between income and traffic offenses. Lastly, the data does not account for citations that were dismissed or dropped after being reviewed, which could affect the accuracy of the results.


\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Linear Regression

In @fig-ppcheckandposteriorvsprior-1 we implement a p	
Simple Linear Regression Model. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: ressidual_plots
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Creates a formula Y = B_O + B_1 X"]

#### Load saved model ####
first_model <- readRDS(here::here("models/sim_offence_linear_regression.rds"))

#### Display model summary ####
summary(first_model)

```

## Diagnostics

@fig-ResidualDiagnostics-1 is a Residual vs. Fitted Values Plot. It shows... This suggests...

@fig-ResidualDiagnostics-2 is a Normal Q-Q Plot. It shows... This suggests...

@fig-ResidualDiagnostics-3 is a . Residuals vs. Predictors Plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-ResidualDiagnostics
#| fig-cap: "Diagnostic plots to assess residuals."
#| fig-subcap: ["Residuals vs. Fitted Values", "Normal Q-Q Plot", "Residuals vs. Predictors"]
#| layout-ncol: 2

#### Residual diagnostics ####
par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid
plot(first_model)  # This will generate residual plots automatically

```
\newpage


# References


